{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-04-29 paper.tex
more edits

\section{The FastKernel method}\label{sec:FastKernel}

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering cross-section. Consider for example the calculation of a general cross-section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation}\label{eq:pertexp}
  \sigma_{pp\to X} =
  \sum_{i,j}\sum_{p=0} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)}_{ij\to X}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)\,.
\end{equation}
where $Q^2$ is the typical hard scale of the process, the indices
$i,j$ sum over the active partonic species in the calculation, $p_{\rm LO}$ is the leading-order power of $\alpha_s$ for the process
and $\hat{\sigma}^{(p)}_{ij\to X}$ the N$^p$LO contribution.

The central observation of tools such as {\tt APPLgrid} and {\tt
  FastNLO} is that the PDF and $\alpha_S$ dependence may be factorised out of
  the convolution via expansion over a set of interpolating functions, spanning $Q^2$ and the
  two values of parton-$x$. For example one may represent the PDFs and $\alpha_S$ as
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,f_i(x_1,Q^2) f_j(x_2,Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)
  \,\mathcal{I}_\tau(Q^2)\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)
\end{array}
\end{equation}
in terms of the Lagrange basis polynomials $\mathcal{I}_\tau(Q^2)$,
$\mathcal{I}_\alpha(x_1)$, $\mathcal{I}_\beta(x_2)$.
Using these expressions for the PDFs in the double-convolution of Eq.~(\ref{eq:pertexp}) one finds:
\begin{equation}\label{eq:XsecPDFs}
\begin{array}{rcl}
  \displaystyle \sigma_{pp\to X} &=& \displaystyle 
                                     \sum_{i,j}\sum_{p=0}^{N_\alpha} \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
                                     \,f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\\
  \\
                                 &\times& \displaystyle \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
                                          \hat{\sigma}^{(p)}_{ij\to X}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
                                          \,\mathcal{I}_\beta^{(n_2)}(x_2)\,.
\end{array}
\end{equation}

Additionally one may take advantages of symmetries present in the matrix element to reduce the sum over active flavours $(i,j)$ in
Eq.~(\ref{eq:hadconv}) to a reduced number of partonic subprocesses with a replacement such as
\begin{equation}
\sum_{i,j} \hat{\sigma}^{(p)}_{ij\to X}
f_i(x_\alpha,Q_\tau^2)\,f_j(x_\beta,Q_\tau^2)\rightarrow
\sum_{s}^{N_{\rm sub}} \hat{\sigma}^{(p)(s)} F^{(s)}_{\alpha\beta,\tau}
\end{equation}
where the $s$ index runs over the $N_{\rm sub}$ (process-specific)
parton subprocess:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}_{\alpha\beta,\tau} =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{\alpha},Q^2_\tau)f_j(x_{\beta},Q^2_\tau)\,,
\end{equation}
and the $C^{(s)}_{ij}$ matrix encodes the combination of PDFs that
contribute to the $s$-th subprocess and $\hat{\sigma}^{(p)(s)}$
is the corresponding subprocess hard cross section. Finally, defining:
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau^{(n')}(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha^{(n_1)}(x_1)
  \,\mathcal{I}_\beta^{(n_2)}(x_2)\,,
\end{equation}
one can write Eq.~(\ref{eq:XsecPDFs}) as follows:
\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}

Additionally, a number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa} or {\tt HOPPET}~\cite{Salam:2008qg},
are available which perform PDF evolution via an analogous
procedure. {\tt APFEL} in particular, is able to provide the so-called
\textit{evolution operator}. The evolution operator, is computed
assuming that the DGLAP evolution can be encoded in a single operator
whose action on the intial scale PDFs proceduces the final scale
PDFs. Taking $Q_0$ as an initial scale and $Q_\tau$ as a final scale,
on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution
equation with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}

>>>>

\section{Interpolation tools for collider observables}\label{sec:FastKernel}

Hadron collider observables are typically computed in QCD by a double
convolution of parton densities with a hard scattering cross-section. Consider for example the calculation of a general cross-section $pp\to X$ with a set of PDFs $\{f\}$:
\begin{equation}\label{eq:hadconv}
  \sigma_{pp\to X} =
  \sum_{s}\sum_{p} \int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)\,.
\end{equation}
where $Q^2$ is the typical hard scale of the process, the index
$s$ sums over the active partonic subprocesses in the calculation, $p_{\rm LO}$ is the leading-order power of $\alpha_s$ for the process
and $\hat{\sigma}^{(p)(s)}$ is the N$^p$LO cross-section for the partonic subprocess scattering $s\to X$. $F^{(s)}$ represents the 
subprocess parton density:
\begin{equation}\label{eq:APPLsubproc}
  F^{(s)}(x_1,x_2, Q^2) =\sum_{i,j} C^{(s)}_{ij} 
  f_i(x_{1},Q^2)f_j(x_{2},Q^2)\,,
\end{equation}
where the $C^{(s)}_{ij}$ matrix enumerates the combinations of PDFs that
contribute to the $s$-th subprocess.
The central observation of tools such as {\tt APPLgrid} and {\tt
  FastNLO} is that the PDF and $\alpha_S$ dependence may be factorised out of
  the convolution via expansion over a set of interpolating functions, spanning $Q^2$ and the
  two values of parton-$x$. For example one may represent the subprocess PDFs and $\alpha_S$ in terms of the Lagrange basis polynomials $\mathcal{I}_\tau(Q^2)$, $\mathcal{I}_\alpha(x_1)$ and $\mathcal{I}_\beta(x_2)$ as:
\begin{equation}\label{eq:interpolation}
\begin{array}{c}
\displaystyle \alpha_s^{p+p_{\rm LO}}(Q^2) \,F^{(s)}(x_1,x_2, Q^2)
  =\\
\\
\displaystyle \sum_{\alpha,\beta,\tau} \alpha_s^{p+p_{\rm LO}}(Q_\tau^2)
  \,F^{(s)}_{\alpha\beta , \tau}
  \,\mathcal{I}_\tau(Q^2)\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)
\end{array}
\end{equation}
where we use the shorthand $F^{(s)}_{\alpha\beta ,\tau} = F^{(s)}(x_\alpha, x_\beta,Q_\tau^2)$.
Using these expressions in the double-convolution of Eq.~(\ref{eq:hadconv}) one can finally obtain an expression for the desired
cross-section which depends upon the subprocess PDFs only though a simple product:

\begin{equation} \label{eq:applconv}
  \sigma_{pp\to X} = \sum_p^{N_\alpha} \sum_{s}^{N_{\mathrm{sub}}} \sum_{\alpha,\beta,\tau} 
  \alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \, F_{\alpha\beta,\tau}^{(s)}\,.
\end{equation}
where
\begin{equation}
  W_{\alpha\beta,\tau}^{(p)(s)} = \mathcal{I}_\tau(Q^2)\int dx_1\,dx_2\,
  \hat{\sigma}^{(p)(s)}\,\mathcal{I}_\alpha(x_1)
  \,\mathcal{I}_\beta(x_2)\,,
\end{equation}
consists of the convolution of the hard cross-section with the interpolating polynomials. This information may be stored in a precomputed look-up table.
The final expression for the cross-section in Eqn.~\ref{eq:applconv} therefore being considerably more computationally efficient inside a fit than the direct
performance of the double-convolution.

\subsection{The FastKernel method}
A number of tools, $e.g.$ {\tt
  APFEL}~\cite{Bertone:2013vaa} or {\tt HOPPET}~\cite{Salam:2008qg},
are available which perform PDF evolution via an analogous
procedure. {\tt APFEL} in particular, is able to provide the so-called
\textit{evolution operator}. The evolution operator, is computed
assuming that the DGLAP evolution can be encoded in a single operator
whose action on the intial scale PDFs proceduces the final scale
PDFs. Taking $Q_0$ as an initial scale and $Q_\tau$ as a final scale,
on the $x$-space grid one can write:
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where the operator $A$ is the solution of the DGLAP evolution
equation with baundary condition:
\begin{equation}
A^\tau_{\alpha\beta, ik} =
\delta_{\alpha\beta}\delta_{ik}\qquad\mbox{for}\qquad Q_\tau = Q_0\,.
\end{equation}
